Covariates analysis of MSBB RNASeq data

### 2016-05-25 code/Rmd/MSSM_RNASeq_cdr.Rmd
Added markdown to process MSSM data based on cdr scores with a mixed effect model

knit2synapse::knitToFolderEntity(file = "./MSSM_RNASeq.Rmd",
                                 parentId ="syn5570248",
                                 entityName = 'MSBB RNASeq')
```

@@

thisFileName <- 'MSSM_RNASeq.Rmd'

@@

# Create diagnosis column (classified into 3 groups ND, MD, SD)
METADATA$Dx = 'ND'
METADATA$Dx[METADATA$CDR == 1 | METADATA$CDR == 2] = 'MD'
METADATA$Dx[METADATA$CDR >2 ] = 'SD'

# Create brain region x diagnosis column
METADATA$BrainRegion.Dx = paste(METADATA$BrainRegion, METADATA$Dx, sep = '.')

# Select covariates of interest
FactorCovariates <- c('RACE', 'SEX', 'BrainRegion.Dx', 'Batch', 'BB')
ContCovariates <- c('PMI', 'AOD', 'RIN', 'RIN2')

@@

```{r dataExplore, fig.width = 10, fig.height = 10}
# RIN
p = list()
p[[1]] = ggplot(METADATA, aes(x = Dx, y = RIN, fill = BrainRegion)) + geom_boxplot()
p[[1]] = p[[1]] + ggtitle('RIN') + theme(legend.position = 'top')

# AgeAtDeath
p[[2]] = ggplot(METADATA, aes(x = Dx, y = as.numeric(AOD), fill = BrainRegion)) + geom_boxplot()
p[[2]] = p[[2]] + ggtitle('AgeOfDeath') + theme(legend.position = 'top')

# Mean Plaque
p[[3]] = ggplot(METADATA, aes(x = Dx, y = as.numeric(PlaqueMean), fill = BrainRegion)) + geom_boxplot()
p[[3]] = p[[3]] + ggtitle('PlaqueMean') + theme(legend.position = 'top')

# NP.1
p[[4]] = ggplot(METADATA, aes(x = Dx, y = as.numeric(NP.1), fill = BrainRegion)) + geom_boxplot()
p[[4]] = p[[4]] + ggtitle('NP.1') + theme(legend.position = 'top')

@@

# Sex
vcd::mosaic(~ SEX + Dx + BrainRegion, data = METADATA)
```

@@

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=BrainRegion.Dx, shape=SEX, size=PMI))
p <- p + theme_bw() + theme(legend.position="top") + scale_shape_manual(values = 1:7)
# p <- p + geom_text(aes(label= ID), size=4, hjust=1)
p

@@

### Normalisation (iterative design)
Since many covariates are correlated, re-normalising COVARIATES with an iterative design matrix. Here BB (Donor Information) is chosen as random effect
```{r iterativeNorm}
# Primary variable of interest
postAdjustCovars = 'BrainRegion.Dx'

@@

tmp <- paste('Using following covariates in the final model:', modelStr)
```
`r tmp`
### Sanity check
```{r residualSigCovars.manual, fig.width=20, fig.height=12}
residualSigCovars[["PC_res"]][[2]]$plotData
```
### Residual calculation
Calculate weighted residuals and add back "BrainRegion.Dx" to the residuals
```{r varsToAddBack}
# Add variable of interest back to the residuals
varsToAddIn = grep("BrainRegion.Dx", colnames(DM1$design), value = T)
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
  VOOM.ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(DM1$design[,varsToAddIn])
```

@@

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=factor(BrainRegion.Dx),size=as.numeric(PMI)))
p <- p + theme_bw() + theme(legend.position="top") 
# p <- p + geom_text(aes(label= ID), size=4, hjust=0)
p

@@

# Query biomart
Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol"),
                       filters = "ensembl_gene_id", values = rownames(RESIDUAL.GENE_EXPRESSION),
                       mart = mart)
RESIDUAL.GENE_EXPRESSION <- left_join(rownameToFirstColumn(RESIDUAL.GENE_EXPRESSION, "ensembl_gene_id"),
                                      Ensemble2HGNC) %>%
  dplyr::select(ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
VOOM.ADJUSTED.GENE_EXPRESSION$E <- left_join(rownameToFirstColumn(VOOM.ADJUSTED.GENE_EXPRESSION$E, "ensembl_gene_id"),
                                             Ensemble2HGNC) %>%
  dplyr::select(ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
```
```{r summary, results='asis', echo=FALSE}
writeLines('Number of samples in each category are')
tmp = as.data.frame(summary(COVARIATES$BrainRegion.Dx))
colnames(tmp) = 'No. of samples'
kable(tmp)
```
### Differential expression analysis
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp, fig.height=10, fig.width=18}
# Fit contrast
contrast = makeContrasts(contrasts=c("BrainRegion.DxBM_10.ND-BrainRegion.DxBM_10.MD", 
                                     "BrainRegion.DxBM_10.ND-BrainRegion.DxBM_10.SD", 
                                     "BrainRegion.DxBM_10.MD-BrainRegion.DxBM_10.SD",
                                     "BrainRegion.DxBM_22.ND-BrainRegion.DxBM_22.MD", 
                                     "BrainRegion.DxBM_22.ND-BrainRegion.DxBM_22.SD", 
                                     "BrainRegion.DxBM_22.MD-BrainRegion.DxBM_22.SD",
                                     "BrainRegion.DxBM_36.ND-BrainRegion.DxBM_36.MD", 
                                     "BrainRegion.DxBM_36.ND-BrainRegion.DxBM_36.SD", 
                                     "BrainRegion.DxBM_36.MD-BrainRegion.DxBM_36.SD",
                                     "BrainRegion.DxBM_44.ND-BrainRegion.DxBM_44.MD", 
                                     "BrainRegion.DxBM_44.ND-BrainRegion.DxBM_44.SD", 
                                     "BrainRegion.DxBM_44.MD-BrainRegion.DxBM_44.SD"), 
                         levels = colnames(VOOM.ADJUSTED.FIT$coefficients))
FIT = contrasts.fit(VOOM.ADJUSTED.FIT, contrasts=contrast)
FIT = eBayes(FIT)

# Get differnetial expression
DE = list()
p = list()
for (i in 1:dim(contrast)[2]){
  DE[[i]] = topTable(FIT, coef=i, number = dim(VOOM.ADJUSTED.GENE_EXPRESSION$E)[1])
  DE[[i]] = rownameToFirstColumn(DE[[i]], "ensembl_gene_id")
  p[[i]] <- ggplot(DE[[i]], aes(x = logFC, y = -log10(adj.P.Val))) + geom_point()
  p[[i]] <- p[[i]] + ggtitle(gsub('BrainRegion.Dx','',colnames(contrast)[i])) + xlim(c(-1.5,1.5)) + ylim(c(0,6))
  p[[i]] <- p[[i]] + geom_hline(yintercept = -log10(0.05), color = 'red')
}
multiplot(plotlist=p, cols = 4)
names(DE) = gsub('BrainRegion.Dx','',colnames(contrast))

@@

kable(tmp)
```

### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
# Code
CODE <- Folder(name = "MSBB RNASeq", parentId = parentId)
CODE <- synStore(CODE)

# Expression
write.table(RESIDUAL.GENE_EXPRESSION, file='expression.tsv', sep = '\t', row.names=F, quote=F)
obj = File('expression.tsv', 
           name = 'Voom Adjusted Weighted Residual Expression (BrainRegion.Dx added)', 
           parentId = CODE$properties$id)
expr_obj = synStore(obj,
                    used = ALL_USED_IDs, 
                    activityName = activityName, 
                    executed = thisFile, 
                    activityDescription = activityDescription)

@@

# Covariates
covar = rownameToFirstColumn(COVARIATES, 'SampleId')
write.table(covar, file='covariates.tsv', sep = '\t', row.names=F, quote=F)
obj = File('covariates.tsv', 
           name = 'Covariates', 
           parentId = CODE$properties$id)
covar_obj = synStore(obj,
               used = ALL_USED_IDs, 
               activityName = activityName, 
               executed = thisFile, 
               activityDescription = activityDescription)

# Store differential expression results
write.table(DE, file = 'diffExpression.tsv', sep = '\t', row.names=F, quote=F)
dexp_obj = File('diffExpression.tsv', name = 'Differential Expression Analysis', parentId = CODE$properties$id)
dexp_obj = synStore(dexp_obj, used = ALL_USED_IDs, activityName = activityName, 
                   executed = thisFile, activityDescription = activityDescription)
```
|  *Results*            |  *SynapseID*                    |
|  -------              |   ---------                     |
|  Covariates           |  `r covar_obj$properties$id`    |
|  Residual Expression  |  `r expr_obj$properties$id`     |
|  Differential Expression  |  `r dexp_obj$properties$id`      |

>>>>

knit2synapse::knitToFolderEntity(file = "./MSSM_RNASeq_cdr.Rmd",
                                 parentId ="syn5570248",
                                 entityName = 'MSBB RNASeq (based on CDR)')
```

@@

thisFileName <- 'MSSM_RNASeq_cdr.Rmd'

@@

# Create brain region x diagnosis column
METADATA$BrainRegion.CDR = paste(METADATA$BrainRegion, METADATA$CDR, sep = '.')

# Select covariates of interest
FactorCovariates <- c('RACE', 'SEX', 'BrainRegion.CDR', 'Batch', 'BB')
ContCovariates <- c('PMI', 'AOD', 'RIN', 'RIN2', 'PlaqueMean')

@@

```{r dataExplore, fig.width = 10, fig.height = 10}
# RIN
p = list()
p[[1]] = ggplot(METADATA, aes(x = CDR, y = RIN, fill = BrainRegion)) + geom_boxplot()
p[[1]] = p[[1]] + ggtitle('RIN') + theme(legend.position = 'top')

# AgeAtDeath
p[[2]] = ggplot(METADATA, aes(x = CDR, y = as.numeric(AOD), fill = BrainRegion)) + geom_boxplot()
p[[2]] = p[[2]] + ggtitle('AgeOfDeath') + theme(legend.position = 'top')

# Mean Plaque
p[[3]] = ggplot(METADATA, aes(x = CDR, y = as.numeric(PlaqueMean), fill = BrainRegion)) + geom_boxplot()
p[[3]] = p[[3]] + ggtitle('PlaqueMean') + theme(legend.position = 'top')

# NP.1
p[[4]] = ggplot(METADATA, aes(x = CDR, y = as.numeric(NP.1), fill = BrainRegion)) + geom_boxplot()
p[[4]] = p[[4]] + ggtitle('NP.1') + theme(legend.position = 'top')

@@

# Sex
vcd::mosaic(~ SEX + CDR + BrainRegion, data = METADATA)
```

@@

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=BrainRegion.CDR, shape=SEX, size=PMI))
p <- p + theme_bw() + theme(legend.position="top") + scale_shape_manual(values = 1:7)
# p <- p + geom_text(aes(label= ID), size=4, hjust=1)
p

@@

### Normalisation (iterative design)
Since many covariates are correlated, re-normalising COVARIATES with an iterative design matrix. Here BB (Donor Information) is chosen as random effect
```{r iterativeNorm}
# Primary variable of interest
postAdjustCovars = 'BrainRegion.CDR'

@@

tmp <- paste('Using following covariates in the final model:', modelStr)
```
`r tmp`
### Sanity check
```{r residualSigCovars.manual, fig.width=20, fig.height=12}
residualSigCovars[["PC_res"]][[2]]$plotData
```
### Residual calculation
Calculate weighted residuals and add back "BrainRegion.CDR" to the residuals
```{r varsToAddBack}
# Add variable of interest back to the residuals
varsToAddIn = grep("BrainRegion.CDR", colnames(DM1$design), value = T)
RESIDUAL.GENE_EXPRESSION = RESIDUAL.GENE_EXPRESSION + 
  VOOM.ADJUSTED.FIT$coefficients[,varsToAddIn] %*% t(DM1$design[,varsToAddIn])
```

@@

p <- ggplot(plotdata, aes(x=PC1, y=PC2))
p <- p + geom_point(aes(color=factor(BrainRegion.CDR),size=as.numeric(PMI)))
p <- p + theme_bw() + theme(legend.position="top") 
# p <- p + geom_text(aes(label= ID), size=4, hjust=0)
p

@@

# Query biomart
Ensemble2HGNC <- getBM(attributes = c("ensembl_gene_id", "hgnc_symbol"),
                       filters = "ensembl_gene_id", values = rownames(RESIDUAL.GENE_EXPRESSION),
                       mart = mart)
RESIDUAL.GENE_EXPRESSION <- left_join(rownameToFirstColumn(RESIDUAL.GENE_EXPRESSION, "ensembl_gene_id"),
                                      Ensemble2HGNC) %>%
  dplyr::select(ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
VOOM.ADJUSTED.GENE_EXPRESSION$E <- left_join(rownameToFirstColumn(VOOM.ADJUSTED.GENE_EXPRESSION$E, "ensembl_gene_id"),
                                             Ensemble2HGNC) %>%
  dplyr::select(ensembl_gene_id, hgnc_symbol, one_of(rownames(COVARIATES)))
```
```{r summary, results='asis', echo=FALSE}
writeLines('Number of samples in each category are')
tmp = as.data.frame(summary(COVARIATES$BrainRegion.CDR))
colnames(tmp) = 'No. of samples'
kable(tmp)
```
### Differential expression analysis
Genes that are differentially expressed at an FDR <= 0.05 are
```{r diffExp, fig.height=10, fig.width=18}
# Fit contrast
contrast = makeContrasts(contrasts = sapply(unique(METADATA$BrainRegion), function(x){
  apply(combn(paste0('BrainRegion.CDR',x,'.',unique(METADATA$CDR)),2), 2, paste, collapse = '-')
}) %>%
  as.character(), 
levels = colnames(VOOM.ADJUSTED.FIT$coefficients))
FIT = contrasts.fit(VOOM.ADJUSTED.FIT, contrasts=contrast)
FIT = eBayes(FIT)

# Get differnetial expression
DE = list()
p = list()
for (i in 1:dim(contrast)[2]){
  DE[[i]] = topTable(FIT, coef=i, number = dim(VOOM.ADJUSTED.GENE_EXPRESSION$E)[1])
  DE[[i]] = rownameToFirstColumn(DE[[i]], "ensembl_gene_id")
  p[[i]] <- ggplot(DE[[i]], aes(x = logFC, y = -log10(adj.P.Val))) + geom_point()
  p[[i]] <- p[[i]] + ggtitle(gsub('BrainRegion.CDR','',colnames(contrast)[i])) + xlim(c(-1.5,1.5)) + ylim(c(0,6))
  p[[i]] <- p[[i]] + geom_hline(yintercept = -log10(0.05), color = 'red')
}
# multiplot(plotlist=p, cols = 4)
names(DE) = gsub('BrainRegion.CDR','',colnames(contrast))

@@

kable(tmp[,1:2])
```

### Store files in synapse
```{r synapse.store, include=FALSE, eval=TRUE, cache=FALSE}
# Code
CODE <- Folder(name = "MSBB RNASeq (based on CDR)", parentId = parentId)
CODE <- synStore(CODE)

# Expression
write.table(RESIDUAL.GENE_EXPRESSION, file='MSSMexpression.tsv', sep = '\t', row.names=F, quote=F)
obj = File('MSSMexpression.tsv', 
           name = 'Voom Adjusted Weighted Residual Expression (BrainRegion.CDR added)', 
           parentId = CODE$properties$id)
expr_obj = synStore(obj,
                    used = ALL_USED_IDs, 
                    activityName = activityName, 
                    executed = thisFile, 
                    activityDescription = activityDescription)

@@

# Covariates
covar = rownameToFirstColumn(COVARIATES, 'SampleId')
write.table(covar, file='MSSMcovariates.tsv', sep = '\t', row.names=F, quote=F)
obj = File('MSSMcovariates.tsv', 
           name = 'Covariates', 
           parentId = CODE$properties$id)
covar_obj = synStore(obj,
               used = ALL_USED_IDs, 
               activityName = activityName, 
               executed = thisFile, 
               activityDescription = activityDescription)

# Store differential expression results
write.table(DE, file = 'MSSMdiffExpression.tsv', sep = '\t', row.names=F, quote=F)
dexp_obj = File('MSSMdiffExpression.tsv', name = 'Differential Expression Analysis', parentId = CODE$properties$id)
dexp_obj = synStore(dexp_obj, used = ALL_USED_IDs, activityName = activityName, 
                   executed = thisFile, activityDescription = activityDescription)
```
|  *Results*                |  *SynapseID*                    |
|  -------                  |   ---------                     |
|  Covariates               |  `r covar_obj$properties$id`    |
|  Residual Expression      |  `r expr_obj$properties$id`     |
|  Differential Expression  |  `r dexp_obj$properties$id`     |
